A Physically Motivated Framework to Compare Merger Timescales\\ of Isolated Low- and High-Mass Galaxies Across Cosmic Time

### 2024-03-17 paper/main.tex
mostly finished the pair sample part of methodology - next up pair sample/calculation specifics

\begin{abstract}
% Statement
Evolutionary differences between low-mass ($\rm 10^8<M_*<5\times10^9\,\Msun$) and high-mass ($\rm 5\times10^9<M_*<10^{11}\,\Msun$) galaxy pairs have been studied in detail for the last few decades, with studies finding differences in star formation rates, gas fractions, and dark matter halo profiles. 
% Problem 
However, many studies continue to use equivalent separation criteria to select galaxy pairs for pair fraction, merger fraction, and merger rate studies, though it is unclear that such a selection permits an equitable comparison between galaxies of different mass, nor as a function of redshift.   
% Our solution
We use the Illustris TNG100 simulation to quantify the merger timescales of isolated low-mass and high-mass galaxy pairs as a function of cosmic time. 
% What are the timescales? 
We show that separation limits that vary with the mass and redshift of the system, such as scaling by the virial radius of the host halo ($r_{\mathrm{sep}}< 1 R_{\rm vir}$), lead to equivalent merger timescales for Low- and High-mass systems at all redshifts. 
Alternatively, static physical separation selections applied equivalently to all galaxy pairs at all redshifts, particularly for the closest pairs ($\rsep<150\kpc$) leads to timescales that differ between low- and high-mass pairs. 
% Why it matters 
Our findings suggest that the comparison of low-mass and high-mass galaxy pair fractions, merger timescales, and merger rates must be done with self-consistent separation criteria that scales as a function of mass and redshift of the system. 
As a result, applying the same merger timescales to different mass systems will lead to a bias that systematically (over/under) predicts low-mass galaxy merger rates. 
\end{abstract}

@@

$z=1.5$ corresponds to snapshot \#40 in the TNG sim
% Introduce RG15, Lotz, etc.

%%% OUTLINE
\section{Sample Data}
%recap last paper and mention which data sample we narrow down to (low mass major pairs only) -- give examples of orbits

\subsection{Isolated Pairs}
We utilize the \paircat{} of isolated low and high mass pairs described in ~\citet{Chamberlain2024}. 
In summary, the \paircat{} is a collection of isolated subhalo pairs at each snapshot of the highest resolution run of the $(110.7\,\Mpc)^3$ volume IllustrisTNG simulation, TNG100-1 (hereafter TNG100).

Stellar masses were assigned to Abundance matching 
For the purposes of our study, we will only be using the

\subsubsection{brief recap of last paper}
\texttt{SUBLINK} algorithm~\citep{RG2015}

\subsubsection{Assumption of Isolation}
% justify stuff from last paper

\subsection{Mergers and Orbits}
\subsubsection{Calculation of merger fraction of total sample}
Plot \#1: Merger fraction either before or after orbits
Note that it is very interesting that dwarf pairs in the field have high merger fractions, while the likelihood of merger is very low in more dense environments (add citation).

\subsubsection{Pulling Orbits}
Orbital data for each pair from the \paircat{} was collected by following each subhalo of the pair both forwards and backwards in time via the \texttt{SUBLINK} algorithm~\citep{RG2015} merger trees.

The infall snapshot is defined as ZZZ. 
We will define the "First Infall" as the first snapshot where the primary and secondary have the same parent FoF halo.

Mergers are XXX 
Non-mergers are YYY. 
The merger snapshot is defined ALALAL.

The orbit collection includes mergers and non-mergers, which

Plot \#2: Orbits plot
Figure~\ref{fig:example-orbits} shows a collection of example orbits pulled from the merger trees of low-mass major pairs initially selected at $z=2$.

 Non-mergers: These orbits can have a variety of orbital periods and thus, a number of different pericentric passages. The two lighter blue orbits are very long period orbits with 1-2 pericenter passages in the past 10Gyr, while the darker blue orbits has a much shorter period with many close encounters throughout, and 3 close passages in just the past 2 Gyr.

Mergers:  Pairs chosen at $z=2$ on average merge at $z=XX$/within XX Gyrs, though the spread of merger timescales is large, leading to a single redshift-selected sample having merger timescales between 0.5-8Gyr.

Flybys:

\subsubsection{Interpolation}
\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/1_fmerge_comp.png}
    \caption{The fraction of pairs that merge before z=0 as a function of the redshift where they were selected via the selection criteria for the \paircat{} from \citet{Chamberlain2024}. At $z>1$, the fraction of selected pairs that merge before $z=0$ is upwards of 80\% for both low mass and high mass isolated pairs. The sharp decline to $F_{\rm merge}$ at $z=0$ is a non-physical feature of the simulation ending at $z=0$, thus unable to follow the pairs after the simulation ends to see what fraction of pairs may have merged.}
    \label{fig:fmerge}
    \end{center}
\end{figure}

@@

@@

% \begin{figure*}[htb]
%     \centering
%     \includegraphics[width=\textwidth]{plots/bet-on-it/3_Timevsz.png}
%     \caption{}
%     % \label{fig:low-vs-high}
% \end{figure*}
% \begin{figure*}[htb]
%     \centering
%     \includegraphics[width=\textwidth]{plots/bet-on-it/3_Timevsz-2d.png}
%     \caption{}
%     % \label{fig:low-vs-high}
% \end{figure*}
\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{plots/bet-on-it/3_time_til_merger_full_fit.png}
    \caption{}
    % \label{fig:low-vs-high}
\end{figure*}

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{plots/bet-on-it/3_time_til_merger.png}
    \caption{}
    % \label{fig:low-vs-high}
\end{figure*}

>>>>

\begin{abstract}
% Statement
Evolutionary differences between low-mass ($\rm 10^8<M_*<5\times10^9\,\Msun$) and high-mass ($\rm 5\times10^9<M_*<10^{11}\,\Msun$) galaxy pairs have been studied in detail for the last few decades, with studies finding differences in star formation rates, gas fractions, and dark matter halo profiles. 
% Problem 
However, many studies continue to use equivalent separation criteria to select galaxy pairs for pair fraction, merger fraction, and merger rate studies, though it is unclear that such a selection permits an equitable comparison between galaxies of different mass, nor as a function of redshift.   
% Our solution
We use the Illustris TNG100 simulation to quantify the merger timescales of isolated low-mass and high-mass major pairs as a function of cosmic time. 
% What are the timescales? 
We show that separation limits that vary with the mass and redshift of the system, such as scaling by the virial radius of the host halo ($r_{\mathrm{sep}}< 1 R_{\rm vir}$), lead to equivalent merger timescales for low- and high-mass systems at all redshifts. 
Alternatively, static physical separation selections applied equivalently to all galaxy pairs at all redshifts, particularly for the closest pairs ($\rsep<150\kpc$) leads to timescales that differ between low- and high-mass pairs. 
% Why it matters 
Our findings suggest that the comparison of low-mass and high-mass galaxy pair fractions, merger timescales, and merger rates must be done with self-consistent separation criteria that scales as a function of mass and redshift of the system. 
As a result, applying the same merger timescales to different mass systems will lead to a bias that systematically (over/under) predicts low-mass galaxy merger rates. 
\end{abstract}

@@

% $z=1.5$ corresponds to snapshot \#40 in the TNG sim
% Introduce RG15, Lotz, etc.

%%% OUTLINE
\section{Methodology}
%recap last paper and mention which data sample we narrow down to (low mass major pairs only) -- give examples of orbits
We utilize the group catalogs, produced by the \texttt{SUBFIND} algorithm~\citep{Springel2001,Dolag2009}, and the merger tree catalogs, generated by the \texttt{SUBLINK} algorithm~\citep{RG2015}, from the highest resolution run of the IllustrisTNG simulation TNG100-1 (hereafter TNG100) to study the merger timescales of galaxy pairs from $z=0-XX$.

Our sample consists of major low-mass and high-mass pairs that are isolated but physically associated.
From this sample, we will determine the fraction of pairs at each snapshot that merge before $z=0$, and track the orbits of the subhalos of each pair to study their merger timescales.

\subsection{Pair orbit sample}
We begin with a subset of the \paircat{} described in ~\citet{Chamberlain2024}, which consists of a collection of isolated subhalo pairs at each snapshot in TNG100. 
A brief version of the selection routine is described here.

At each snapshot, low-mass and high-mass pairs are chosen by first selecting the two most massive halos (by stellar mass) from FoF groups with virial mass\footnote{We use \texttt{Group\_M\_TopHat200} from the TNG100 Group Catalogs as the FoF Group virial mass. This mass is defined to be the mass enclosed by a sphere with mean density $\Delta_c *\rho_c$, where $\Delta_c$ is the overdensity constant from~\citet{Brynorman1998} and $\rho_c$ is the critical density of the universe at the time calculated. The corresponding virial radius in TNG100 is given by \texttt{Group\_R\_TopHat200}.} 
\begin{align*}
        \mbox{\textbf{low mass:}}&\,\rm M_{G} = 8\times 10^{10}- 5\times 10^{11}\,\Msun \\ 
        \mbox{\textbf{high mass:}}&\, \rm M_{G}=10^{12}- 6.5\times10^{12}\,\Msun.
\end{align*}
Utilizing the most massive subhalos from the same FoF group ensures that the pairs are isolated from massive nearby systems that may perturb the dynamical state of the pair.

We require that subhalos that constitute a pair meet a minimum subhalo mass criteria of 
\begin{equation*}
    \mbox{\textbf{minimum subhalo mass:}}\,
    \Mhalo > 1\times10^{9}\Msun.
\end{equation*}
at the snapshot of consideration. 
For each halo in the FoF group that passes the minimum subhalo mass criteria, we utilize the \texttt{SUBLINK} catalogs to trace their mass history~\citep{RG2015}, and find the peak halo mass of each subhalo.

Stellar masses are then assigned to each subhalo in the FoF group using the Abundance Matching prescription of \citet{Moster2013}. 
The peak halo mass and current snapshot are used to calculate the stellar mass of each subhalo, so that the abundance matching prescription is of the form $\ms{}=f(\Mpeak,z_{\mathrm{snap}})$.

In \citet{Chamberlain2024}, the abundance matching prescription is sampled 1000 times for each subhalo to account for the spread in the abundance matching prescription. 
However, for the purposes of the present study, we only use the stellar masses given by the median of the abundance matching relationship. 
This choice allows us to more thoroughly consider the population under study, and also permits a roughly 1:1 monotonic relationship between stellar mass and peak halo mass, so that taking subsamples of the lowest stellar mass objects will result in a subsample with lower peak halo mass as well. % is this a good enough reason? 
% else, maybe spread of number of pairs is so small, that we expect the noise of the timescale analysis to be dominated by the snapshot spacing anyways?

Primary halos are defined as the subhalo with the highest stellar mass ($M_{*1}$) in their FoF group, and secondaries are defined as the second most massive subhalo with stellar mass $M_{*2}$. 
Our sample of major pairs then consists of all pairs of primary and secondary halos with 
\begin{align*} 
\mbox{\textbf{low mass primaries:}}&\, 10^{8}< \rm M_{*1} < 5\times10^{9} \Msun \\ 
\mbox{\textbf{high mass primaries:}}&\, 5\times 10^{9}< \rm M_{*1} < 10^{11} \Msun\\
\mbox{\textbf{stellar mass ratio:}}&\,      
    M_{*2}/M_{*1} > 1/4.
\end{align*}
A primary or secondary halo can only be a member of one single pair at a given snapshot, such that a collection of $N$ pairs consists of $N$ unique primaries and $N$ unique secondaries.

% i am removing the 10kpc component of this because I think it'd just be easier to handle those cases in the 
% Finally, we will only consider pairs that have physical separations of $\rsep>10\,\kpc$ at the snapshot of selection. 
% The physical separation cut does not vary with redshift, and is the same for both low-mass and high-mass pairs.

The base sample used for this analysis then consists of the set of all low-mass and high-mass major pairs from each snapshot of the TNG100 simulation.

\subsubsection{Mergers}
% Plot \#1: Merger fraction either before or after orbits
% Note that it is very interesting that dwarf pairs in the field have high merger fractions, while the likelihood of merger is very low in more dense environments (add citation).

The primary and secondary subhalos of each pair have a `SubfindID' that is used to identify them in the \texttt{Subfind} catalogs, which contain information about every FoF group and subhalo at each snapshot. 
The \texttt{SUBLINK} catalogs track subhalos from one snapshot to the next, and thus enable us to track subhalos both backwards and forwards in time from any given snapshot.

We utilize the \texttt{RootDescendantID} and \texttt{DescendantID} fields of the \texttt{SUBLINK} catalogs to determine which pairs from our base sample merge before the end of the simulation (at $z=0$), and to determine when the merger occurs. 
The \texttt{DescendantID} field provides the `SubhaloID'\footnote{Note that the `SubhaloID' is distinct from the `SubfindID', and is unique for every subhalo in the merger trees. 
The \texttt{SUBLINK} catalogs provide the associated `SubfindID' of each subhalo.} of the subhalo's descendent in the next (or one of the following) snapshot, if it has one. 
% By following the \texttt{DescendantID}, you follow a subhalo through all of it's mergers to the `root' of the merger tree.  
The \texttt{RootDescendantID} field provides the SubhaloID of the root subhalo, which is the latest subhalo in the simulation that is a descendent of the subhalo.

We define the set of merging pairs as the collection of all pairs where the primary and secondary halo have the same \texttt{RootDescendantID}, meaning that the branches of their merger trees have the same `root', or common descendent. 
If a given pair has two distinct \texttt{RootDescendantID} values, the pair is defined as a `non-merger.'

For each merging pair, we define the `merger snapshot' as the snapshot that immediately follows the first snapshot where the primary and secondary have the same \texttt{DescendantID}. 
For example, if the primary and secondary have different \texttt{DescendantIDs} for snapshots 40-45, but have the same \texttt{DescendantID} at snapshot 46, then the merger must take place some time between snapshot 46 and 47. 
We take snapshot 47 to be the merger snapshot.

\subsubsection{Orbits} 
We calculate orbits for all mergers and non-mergers in our base sample. 
An orbit for a single pair is defined to be the physical separation between the primary and secondary halo as a function of redshift or Lookback Time.

A given pair from the base sample at snapshot $S_n$, which by definition passes all selection criteria at $S_n$, can be followed backwards and forwards in time using the \texttt{SUBLINK} merger trees.  
We track the positions of both the primary and secondary subhalo at each snapshot and calculate the physical separations after accounting for the periodic boundary conditions of the simulation box.
In cases where the primary or secondary does not have a defined position at a given snapshot, we set the separation at that snapshot to NaN.\footnote{If a subhalo is very small, or is passing through a more massive subhalo, and is unable to reach the density contrast required to be identified as an independent structure by the Subfind algorithm, it will not have a defined position in the Sublink catalogs. The sublink algorithm allows for subhalos to skip a single snapshot, and identifies the `skipped descendent' in the $S_{n+2}$ snapshot, so that the orbit can be computed contiguously for a pair on . See Sec.~3 in ~\citet{RG2015} for more details.}

In addition to the physical separation, we also calculate the scaled separation of a pair at each snapshot. We use the definition of scaled separation from~\cite{Chamberlain2024}, given as
\begin{equation}
    \rm \Rsc = \frac{\rsep}{\Rvir}
\end{equation}
where $\rsep$ is the physical separation in $\kpc$, and $\Rvir$ is the virial radius of the primary's FoF group, which reasonably approximates the virial radius of a halo with a virial mass equal to the combined subhalo mass of the primary and secondary.
The scaled radius $\Rsc$ is by construction a function of mass and redshift, which will account for the mass difference between low-mass and high-mass pairs, or for halo growth over time.

We collect the following data for each pair at each snapshot where the primary and secondary are both defined: the virial mass of the primary's FoF group \texttt{Group\_M\_TopHat200}, the virial radius of the primary's FoF group \texttt{Group\_R\_TopHat200}, the physical separation (in $\kpc$), and the scaled separation (dimensionless).

\subsubsection{"First Infall"}
The orbit of a pair may be technically defined at very early times, well before the primary and secondary are physically associated. 
However, we wish to constrain our orbital analysis to only physically associated pairs, and thus will not consider the orbit of a subhalo pair before they are a part of the same FoF group. Specifically, we will consider only the "Post-Infall" part of each pair's orbit.

We define the snapshot of "First Infall" as the first snapshot where the primary and secondary have the same parent FoF halo, and "Post-Infall" as the set of snapshots including the snapshot of "First Infall" and all snapshots thereafter. 
The post-infall orbit definition is slightly more robust for considering the full interaction timescales of merging pairs than only considering the orbit when the FoF group of the primary and secondary are the same. 
This is because there are cases of pair interactions where the secondary will enter the FoF group of the primary, have a pericentric passage, then leave the FoF group for a few gigayears before re-entering the primary FoF group and eventually merging with the primary.

An example of the orbit of a merging pair where a "common group" criteria may differ from a "post-infall" can be seen in Figure~\ref{fig:example-orbits}, which shows the variety of orbits that can be found in the TNG100 simulation for pairs that were originally selected at $z=1.5$. 
The top panel shows the orbits of 5 pairs that merge, where solid lines show the parts of the orbit that occur while the primary and secondary share a common FoF group, and dashed lines show the parts of the merger where the FoF group of the secondary is different than that of the primary. 
The "post-infall" part of the orbit occurs after the first snapshot where the solid lines begins.
Pairs chosen at $z=1.5$ on average merge at $z=XX$/within XX Gyrs, though the spread of merger timescales is large, leading to a single redshift-selected sample having merger timescales between $\sim1-10\Gyr$.

The middle panel shows the orbits for 3 pairs that are classified as non-mergers, since they do not merge before $z=0$, but which appear likely to merge within a couple Gyr past the end of the simulation, and are thus "Potential Future Mergers". 
These orbits can have a variety of orbital periods, and the number of pericentric passages can vary significantly. 
The two lighter blue orbits are very long period orbits with 1-2 pericenter passages in the past 10Gyr, while the darker blue orbit has a much shorter period with many close encounters throughout, and 3 close passages in just the past 2 Gyr.
The bottom panel shows the orbits of "Fly-by" interactions (non-mergers) that are unlikely to merge in the near future, if ever.  
Note that we do not split our non-merger category into fly-bys and potential future mergers for any of our following analyses, and such distinctions were made only for the purposes of showing the diversity of orbits that pairs selected at the same snapshot may follow.

@@

\subsubsection{Uniqueness of pairs}
Since our entire pair sample consists of pairs chosen at separate snapshots, and since a single pair could pass our selection criteria at many snapshots, we keep only a single orbital instance of each pair.

To distinguish pairs from one snapshot to another, each pair is assigned a pairkey while constructing their orbit, which is created by combining the primary's and secondary's earliest SubhaloID from the Sublink catalogs, which will be unique for each pair of halos. 
After each pair is assigned a unique pairkey, we save the orbit as a function of time only once for a pair that otherwise would have been double/multi-counted in our orbit catalog.

The total number of low-mass pairs selected at each snapshot (given by $\Sigma_{S_n} n_p$) the sum of the number of pairs that pass the selection criteria at each snapshot is XXX, while the total number of \textit{unique} pairs is XXX. 
The total number of high-mass pairs is XXX, while the total number of unique pairs is XXX.

Note that a single subhalo may be a member of many different pairs, but will only have one unique orbit per unique pair.
For example, the primary of a low-mass pair selected at $z=3$ and merges before $z=2$ may be selected at a later snapshot with a new secondary companion. 
In this case, both orbits (of the original low-mass pair and the new pair which includes the remnant subhalo from the previous merger) are kept, since having the primary occur twice in the catalog does not mean that the orbit is being double counted ugh.

\subsection{Calculating the Merger Fraction}
\subsection{Calculating the Time Until Merger}

\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/1_fmerge_comp.png}
    \caption{The fraction of pairs that merge before z=0 as a function of the redshift where they were selected via the selection criteria for the \paircat{} from \citet{Chamberlain2024}. At $z>1$, the fraction of selected pairs that merge before $z=0$ is upwards of 80\% for both low mass and high mass isolated pairs. The sharp decline to $F_{\rm merge}$ at $z=0$ is a non-physical feature of the simulation ending at $z=0$, thus unable to follow the pairs after the simulation ends to see what fraction of pairs may have merged.}
    \label{fig:fmerge}
    \end{center}
\end{figure}

\section{Pair Sample Properties}

@@

% \begin{figure*}[htb]
%     \centering
%     \includegraphics[width=\textwidth]{plots/bet-on-it/3_Timevsz.png}
%     \caption{}
%     % \label{fig:low-vs-high}
% \end{figure*}
% \begin{figure*}[htb]
%     \centering
%     \includegraphics[width=\textwidth]{plots/bet-on-it/3_Timevsz-2d.png}
%     \caption{}
%     % \label{fig:low-vs-high}
% \end{figure*}
\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{plots/bet-on-it/3_time_til_merger_full_fit.png}
    \caption{\kc{TODO: add LookbackTime vs. Z line, add Hubble time line} The median time until merger as a function of redshift for low-mass (green) and high-mass (pink) pairs. Shaded regions represent the 1st and 3rd quartile spread on the median. Pairs at each redshift are selected if they have experienced first infall. 
    % 
    The black dashed line shows the Hubble time, which bounds the median time until merger from $z=6$ to $z\sim0.5-0.75$.
    The dotted grey line shows the time to $z=0$ (the Lookback Time) as a function of redshift, which sets the upper bound for the time until merger that a merging pair can have. 
    The median time until merger is similar for low-mass and high-mass pairs, and rises from $z=6$ to a peak at $z\sim0.75$, then decreases to $z=0$.}
    \label{fig:timescales}
\end{figure*}

\begin{figure*}[htb]
    \centering
    \includegraphics[width=\textwidth]{plots/bet-on-it/3_time_til_merger.png}
    \caption{(Top) The median time until merger as a function of redshift for low-mass (green) and high-mass (pink) pairs with 3D physical separations greater than $10\,\kpc$ and less than $(50,70,100,150,200,\mbox{and }300)\,\kpc$ from left to right. 
    (Bottom) The median time until merger as a function of redshift for pairs with physical separations greater than $10\,\kpc$ and less than $(0.25, 0.5, 1, 1.5, 2,\mbox{and }2.5)\,\Rvir$ from left to right. 
    % 
    The time until merger increases from $z=4$ to $z\sim0.5-0.75$, at which point the time until merger decreases to zero since all low redshift mergers must have short merger timescales to merge before the end of the simulation at $z=0$. 
    }
    \label{fig:timescales-sep}
\end{figure*}
